Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2008-06-20 SCP2008/paper/metamonads.tex
all the changes that Oleg recommended

\begin{code}
let simplest_param_code plus one =
  let gen x y = plus x (plus y one) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}
and use it to generate code that operates on integers, floating point
numbers or booleans -- in general, any domain that implements |plus|
and |one|:
\begin{code}
let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
  simplest_param_code plus one
let plus x y = .<.~x || .~y>. and one = .<true>. in
  simplest_param_code plus one
\end{code}
Running the former expression yields a function on |float|s, whereas
the latter expression is a code expression for a boolean function.
This simple technique clearly shows how we can be polymorphic over
domain operations and yet still generate efficient domain-specific code,
thus achieving a proper separation of concerns.

@@

\subsection{Generating control statements}\label{ssec:control}
We can write operations that generate code other than let-statements,
e.g., conditionals: see |ifL| in Figure~\ref{ourmonad}. The function |ifL|, 
albeit straightforward, is not as general as we wish: its arguments are
pieces of code rather than monadic values. We can ``lift it'':
\begin{code}
let ifM' test th el = perform
  testc <-- test; thc <-- th; elc <-- el;
  ifL testc thc elc
\end{code}
However we also need another |ifM| function, with the same
interface (see Figure~\ref{ourmonad}).
The difference between them is apparent from the following example:
\begin{code}[commandchars=\\\{\}]
let gen a i = ifM' (ret .<(.~i) >= 0>.) 
                   (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
\evalresult{.<fun a_1 i_2 ->}
\textcolor{red}{      let t_3 = (Some a_1.(i_2)) in if (i_2 >= 0) then t_3 else None>.}
let gen a i = ifM (ret .<(.~i) >= 0>.) 
                  (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
\evalresult{.<fun a_1 i_2 ->}
\textcolor{red}{      if (i_2 >= 0) then let t_3 = (Some a_1.(i_2)) in t_3 else None>.}
\end{code}
%
If we use |ifM'| to generate guarded array access code, the let-insertion
happens \emph{before} the if-expression, that is, before the test that
the index |i| is positive. If |i| turned out
negative, |a.(i)| would generate an out-of-bound array access
error. On the other hand, the code with |ifM| accesses the array only
after we have verified that the index is non-negative. This example
should make it clear that code generation (such as the one in |retN|) is 
truly an effect, and that we have to be clear about the sequencing of
effects when generating control constructions such as conditionals.
The form |ifM| handles such effects correctly.

@@

We therefore chose a different way of maintaining generator state,
using a monad. We already saw that for let-insertions we could use
a continuation monad;  we now show how to use the state component
of our monad (Fig.~\ref{ourmonad}). The monadic actions |fetch| and
|store| are used to access that monadic state, which is
threaded throughout the entire code-generation computation.

@@

let orec_store ((inj,_,name) as ip:(('a,'b) open_rec)) (v:'a) (s:'b list) 
   : 'b list =
  let () = 
    try let _ = lookup ip s in 
        failwith ("Field "^name^" of an open record is already present."
    with Not_found -> () in
  (inj v)::s

@@

We describe will the following aspects, generally in
order of increasing complexity: domain (of the group `Abstract
domain', see \S\ref{sec:aspect-groups}), code generation combinators
(of the group `Interpretation'), matrix representation (of the group
`Concrete representation'), determinant (the group `Tracking'), and
output (the group `Interface').

@@

The type of our containers includes the lifted domain |Dom| as one of
the components. This is quite convenient since operations on
containers are usually accompanied by operations on retrieved values,
which are subsequently stored again.
The particular instances of the containers are
parametric over a |DOMAINL|, i.e. functors from a |DOMAINL| module to
the actual implementation of a container.  For example, the following
functor defines a matrix container as a single array, with elements
stored in row-major order -- the container used in
Fig.~\ref{fig:GE-algorithm}.
\begin{code}
module GenericVectorContainer(Dom:DOMAINL) =
  struct
  module Dom = Dom
  type contr = Dom.v container2dfromvector
  type 'a vc = ('a,contr) code
  type 'a vo = ('a,Dom.v) code
  let getL x i j = .< ((.~x).arr).(.~i* (.~x).m + .~j) >.
  let dim2 x = .< (.~x).n >.      (* number of rows *)
  let dim1 x = .< (.~x).m >.      (* number of cols *)
  \dots
\end{code}
The accompanying code \cite{metamonadsURL} includes an implementation with
elements stored in a 1D array in a column-wise (Fortran-like)
mode, and another for a matrix represented as an array of rows.

@@

Our sample Gaussian Elimination algorithm (Fig.~\ref{fig:GE-algorithm}),
demonstrates that tracking the determinant is quite complex: first we define
the variables |det_sign| and |det_magn| used for tracking (lines 22 and
23), then we have to change the sign when swapping two rows or two
columns or when the matrix found to be singular (lines 32, 37, 58).
The value of the determinant should be updated for each pivoting (line
55). Finally, we convert the tracking state to the resulting
determinant value (lines 63-65). Most importantly, we observe that these lines
are not contiguous: the determinant aspect is consulted at several separate
places in the GE algorithm, and the aspect is supposed to keep
state. As with other aspects, the determinant aspect is a module,
and has the following signature:
\begin{code}
module type DETERMINANT = sig
  type tdet = C.Dom.v ref
  type 'a lstate
  type 'pc pc_constraint = unit
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  type ('pc,'v) lm = ('pc,'v) cmonad
    constraint _  = 'pc pc_constraint
  type ('pc,'v) om = ('pc,'v) omonad
    constraint _  = 'pc pc_constraint
  type 'pc nm = ('p,unit) monad
    constraint _ = ('pc,'p,_) cmonad_constraint
    constraint _ = 'pc pc_constraint
  val decl : unit -> 'b nm (* no code is generated *)
  val upd_sign  : unit -> ('b,unit) om
  val zero_sign : unit -> ('b,unit) lm
  val acc_magn  : ('a,C.Dom.v) rep -> (<classif : 'a; ..>,unit) lm
  val get_magn  : unit -> ('b,tdet) lm
  val set_magn  : ('a,C.Dom.v) rep -> (<classif : 'a; ..>,unit) lm
  val fin       : unit -> ('b,C.Dom.v) lm
end
\end{code}
The type |'a lstate| denotes the state being kept as part of the
overall monadic state (tagged as |`TDet|). The signature specifies the
operations of the aspect, such as |decl| for initializing the tracking
state, |zero_sign| to record matrix singularity, |set_magn| to set the
magnitude and |fin| to convert the tracking state to the resulting
value, of  the type |C.Dom.v|, the type of the container
elements. Other aspects follow a similar outline.
%\oleg{The following is perhaps too particular to mention in the
%  paper. }
% \jacques{Actually, I liked it, so after slight editing, I kept it}
It is instructive the examine the difference in the return types of the
\texttt{decl}, \texttt{upd\_sign} and \texttt{zero\_sign} functions. The first
says that \texttt{decl} is an action which is executed only for its
side-effect. It yields no interpretation object (e.g., produces no
code value). The function \texttt{zero\_sign} always produces a code
value -- an expression such as assignment that has type unit. 
The function \texttt{upd\_sign} is a generator that may produce code, or may
not. The option type lets us avoid generating pointless code such as 
\texttt{a := b; ()}.

>>>>

\begin{code}
let simplest_param_code plus one =
  let gen x y = plus x (plus y one) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}
and use it to generate code that operates on integers, floating point
numbers or booleans -- in general, any domain that implements |plus|
and |one|:
\begin{code}
let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
  simplest_param_code plus one
let plus x y = .<.~x || .~y>. and one = .<true>. in
  simplest_param_code plus one
\end{code}
Running the former expression yields a function on |float|s, whereas
the latter expression is a code expression for a boolean function.
This simple technique clearly shows how we can abstract over
domain operations and yet still generate efficient domain-specific code,
thus achieving a proper separation of concerns.

@@

\subsection{Generating control statements}\label{ssec:control}
We can write operations that generate code other than let-statements,
e.g., conditionals: see |ifL| in Figure~\ref{ourmonad}. The function |ifL|, 
albeit straightforward, is not as general as we wish: its arguments are
pieces of code rather than monadic values. We can ``lift it'':
\begin{code}
let ifM' test th el = perform
  testc <-- test; thc <-- th; elc <-- el;
  ifL testc thc elc
\end{code}
However we also need another |ifM| function, with the same
interface (see Figure~\ref{ourmonad}).
The difference between them is apparent from the following example:
\begin{code}[commandchars=\\\{\}]
let gen a i = ifM' (ret .<(.~i) >= 0>.) 
                   (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
\evalresult{.<fun a_1 i_2 ->}
\textcolor{red}{      let t_3 = (Some a_1.(i_2)) in if (i_2 >= 0) then t_3 else None>.}
let gen a i = ifM (ret .<(.~i) >= 0>.) 
                  (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
\evalresult{.<fun a_1 i_2 ->}
\textcolor{red}{      if (i_2 >= 0) then let t_3 = (Some a_1.(i_2)) in t_3 else None>.}
\end{code}
%
If we use |ifM'| to generate guarded array access code, the let-insertion
happens \emph{before} the if-expression, that is, before the test that
the index |i| is positive. If |i| turned out
negative, |a.(i)| would generate an out-of-bound array access
error. On the other hand, the code with |ifM| accesses the array only
after we have verified that the index is non-negative. This example
demonstrates that code generation (such as the one in |retN|) is 
truly an effect, and that we have to be clear about the sequencing of
effects when generating control constructions such as conditionals.
The form |ifM| handles such effects correctly.

@@

We therefore chose a different way of maintaining generator state,
using a monad. We already saw that for let-insertions we could use
a continuation monad;  we now demonstrate the state component
of our monad (Fig.~\ref{ourmonad}). The monadic actions |fetch| and
|store| are used to access that monadic state, which is
threaded throughout the entire code-generation computation.

@@

let orec_store ((inj,_,name) as ip:(('a,'b) open_rec)) (v:'a) (s:'b list) 
   : 'b list =
  let () = 
    try let _ = lookup ip s in 
        failwith ("Field "^name^" of an open record is already present.")
    with Not_found -> () in
  (inj v)::s

@@

We will describe the following aspects, generally in
order of increasing complexity: domain (of the group `Abstract
domain', see \S\ref{sec:aspect-groups}), code generation combinators
(of the group `Interpretation'), matrix representation (of the group
`Concrete representation'), determinant (the group `Tracking'), and
output (the group `Interface').

@@

The type of our containers includes the lifted domain |Dom| as one of
the components. This is quite convenient since operations on
containers are usually accompanied by operations on retrieved values,
which are subsequently stored again.
The particular instances of the containers are
parametric over a |DOMAINL|, i.e. functors from a |DOMAINL| module to
the actual implementation of a container.  For example, the following
functor defines a matrix container as a single array, with elements
stored in row-major order -- the container used in
Fig.~\ref{fig:GE-algorithm}.
\begin{code}[commandchars=\\\{\}]
module GenericVectorContainer(Dom:DOMAINL) =
  struct
  module Dom = Dom
  type contr = Dom.v container2dfromvector
  type 'a vc = ('a,contr) code
  type 'a vo = ('a,Dom.v) code
  let getL x i j = .< ((.~x).arr).(.~i* (.~x).m + .~j) >.
  let dim2 x = .< (.~x).n >.      (* number of rows *)
  let dim1 x = .< (.~x).m >.      (* number of cols *)
  \dots
\end{code}
The accompanying code \cite{metamonadsURL} includes an implementation with
elements stored in a 1D array in a column-wise (Fortran-like)
mode, and another for a matrix represented as an array of rows.

@@

Our sample Gaussian Elimination algorithm (Fig.~\ref{fig:GE-algorithm}),
demonstrates that tracking the determinant is quite complex: first we define
the variables |det_sign| and |det_magn| used for tracking (lines 22 and
23), then we have to change the sign when swapping two rows or two
columns or when the matrix found to be singular (lines 32, 37, 58).
The value of the determinant should be updated for each pivoting (line
55). Finally, we convert the tracking state to the resulting
determinant value (lines 63-65). Most importantly, we observe that these lines
are not contiguous: the determinant aspect is consulted at several separate
places in the GE algorithm, and the aspect is supposed to keep
state. As with other aspects, the determinant aspect is a module,
and has the following signature:
\begin{code}
module type DETERMINANT = sig
  type tdet = C.Dom.v ref
  type 'a lstate
  type 'pc pc_constraint = unit
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  type ('pc,'v) lm = ('pc,'v) cmonad
    constraint _  = 'pc pc_constraint
  type ('pc,'v) om = ('pc,'v) omonad
    constraint _  = 'pc pc_constraint
  type 'pc nm = ('p,unit) monad
    constraint _ = ('pc,'p,_) cmonad_constraint
    constraint _ = 'pc pc_constraint
  val decl : unit -> 'b nm (* no code is generated *)
  val upd_sign  : unit -> ('b,unit) om
  val zero_sign : unit -> ('b,unit) lm
  val acc_magn  : ('a,C.Dom.v) rep -> (<classif : 'a; ..>,unit) lm
  val get_magn  : unit -> ('b,tdet) lm
  val set_magn  : ('a,C.Dom.v) rep -> (<classif : 'a; ..>,unit) lm
  val fin       : unit -> ('b,C.Dom.v) lm
end
\end{code}
The type |'a lstate| denotes the state being kept as part of the
overall monadic state (tagged as |`TDet|). The signature specifies the
operations of the aspect, such as |decl| for initializing the tracking
state, |zero_sign| to record matrix singularity, |set_magn| to set the
magnitude and |fin| to convert the tracking state to the resulting
value, of  the type |C.Dom.v|, the type of the container
elements. Other aspects follow a similar outline.
%\oleg{The following is perhaps too particular to mention in the
%  paper. }
% \jacques{Actually, I liked it, so after slight editing, I kept it}
It is instructive to examine the difference in the return types of the
\texttt{decl}, \texttt{upd\_sign} and \texttt{zero\_sign} functions. The first
says that \texttt{decl} is an action which is executed only for its
side-effect. It yields no interpretation object (e.g., produces no
code value). The function \texttt{zero\_sign} always produces a code
value -- an expression such as assignment that has type unit. 
The function \texttt{upd\_sign} is a generator that may produce code, or may
not. The option type lets us avoid generating code such as 
\texttt{a := b; ()} with a pointless |()|.
